---
title: "10x Hudeca Nasal: Integration"
author: "Mbouamboua Yvon <yvon.mbouamboua@inserm.fr>"
output:
    html_document:
       toc: true
       toc_float: true
       number_sections: true
date: 'Last update: `r format(Sys.time())`'
---

# Goal
Merge datasets from different samples while correcting for batch effects.

# Steps
1. Identify batch effects using exploratory data analysis.
2. Apply batch correction methods (Harmony, Seurat CCA, etc.).
3. Assess integration effectiveness using clustering and visualization.

# Setup

```{r setup, include=TRUE, message=FALSE, warning=FALSE}

invisible(gc())
knitr::opts_chunk$set(warning = FALSE, results = TRUE, message = FALSE, comment = "")
dir.main <- "/Users/yvon.mbouamboua/Documents/projects/singlecell/hudeca_nasal"
dir.results <- file.path(dir.main, "results", "002_integration")
dir.create(dir.results, showWarnings = FALSE, recursive = TRUE)

# Load R packages
packages <- c("Seurat", "ggplot2", "dplyr", "tidyverse","grid",
              "SingleCellExperiment", "scran", "scater", "pheatmap", "ComplexHeatmap", "ComplexUpset", "patchwork")

# Function to check and install packages from CRAN or Bioconductor
install_and_load <- function(pkg) {
  # Check if the package is from Bioconductor
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
    BiocManager::install(pkg, ask = FALSE)
  }
  # Load the package
  suppressWarnings(suppressMessages(library(pkg, character.only = TRUE)))
}

# Apply function to each package
suppressPackageStartupMessages({
  invisible(lapply(packages, install_and_load))
})

# Load helper functions
# Define the directory where utility scripts are located
utils_dir <- "~/Documents/projects/singlecell/utilities/"
# Get a list of all R files in the directory
r_files <- list.files(utils_dir, pattern = "\\.R$", full.names = TRUE)
# Source each R file
invisible(suppressWarnings(suppressMessages(lapply(r_files, source))))

```

# Load seurat processed samples

```{r}

seurat.list <- readRDS(file.path(dir.main, "results/001_preprocessing/rdata/seurat.list.rds"))

```


# Integration


```{r fig.width=10, fig.height=6}

seurat.list <- lapply(X = seurat.list, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

invisible(gc())
features <- SelectIntegrationFeatures(object.list = seurat.list)
seurat.list <- lapply(X = seurat.list, FUN = function(x) {
    x <- ScaleData(x, features = features, verbose = FALSE)
    x <- RunPCA(x, features = features, verbose = FALSE)
})

invisible(gc())
anchors <- FindIntegrationAnchors(object.list = seurat.list, anchor.features = features, reduction = "rpca")
obj <- IntegrateData(anchorset = anchors)
rm(anchors)
DefaultAssay(obj) <- "integrated"
obj <- ScaleData(obj, verbose = FALSE)
obj <- RunPCA(obj, npcs = 30, verbose = FALSE)
ElbowPlot(obj)
obj <- RunUMAP(obj, reduction = "pca", dims = 1:30)
obj <- FindNeighbors(obj, reduction = "pca", dims = 1:30)
obj <- FindClusters(obj, resolution = c(0.5,1,2))

saveRDS(obj, file.path(dir.results, "rdata/fn_hudeca_integrated.rds"))

Dim_Plot(object = obj, group.by = "seurat_clusters", label.show = T, legend = T, theme.bw = F,figure.plot = T, legend.ncol = 2)
Dim_Plot(object = obj, group.by = "sample", label.show = F, legend = T, theme.bw = F,figure.plot = T)

```


# Post-processing

## Find doublets

```{r}

obj <- readRDS(file.path(dir.results, "rdata/fn_hudeca_integrated.rds"))

DefaultAssay(obj) <- "integrated"
obj <- FindVariableFeatures(obj, assay = "RNA", nfeatures = 2000)
features <- VariableFeatures(obj, assay = "RNA")
obj <- ScaleData(obj)
obj <- RunPCA(obj, npcs = 30, features = features)
obj <- RunUMAP(obj, reduction = "pca", dims = 1:30, min.dist = 1)
obj <- FindNeighbors(obj, reduction = "pca", dims = 1:30)
#obj <- FindClusters(obj, resolution = c(1))
DefaultAssay(obj) <- "RNA"

sce <- scDblFinder::scDblFinder(as.SingleCellExperiment(obj))
obj$scDblFinder.class <- sce$scDblFinder.class
obj$scDblFinder.class <- unname(obj$scDblFinder.class == "doublet")
obj@meta.data$scDblFinder.class <- as.character(obj@meta.data$scDblFinder.class)
saveRDS(obj, file.path(dir.results, "rdata/fn_hudeca_integrated.rds"))

Dim_Plot(object = obj, group.by = "seurat_clusters", label.show = T, legend = F)
Dim_Plot(object = obj, group.by = "sample")
Dim_Plot(object = obj, group.by = "scDblFinder.class", label.show = F, label.size = 5, legend = T, figure.plot = T)

```


## Filter cell quality

### Check doublets

```{r fig.width=5, fig.height=4}

p <- Dim_Plot(object = obj, group.by = "scDblFinder.class", label.show = F, label.size = 5, legend = T, 
         dark.theme = F, figure.plot = T, legend.position = "top", cell.countours = F, colors.use = c("gray50","darkred"))
p
save_plot(p, filename = file.path(dir.results, "figures/umap_scDblFinder"),  width = 5, height = 4, formats = c("pdf","png"))

```


```{r fig.width=10, fig.height=4, warning=F, message=F}

p <- Stacked_VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mito"), group.by = "seurat_clusters", 
                split.by = "scDblFinder.class", angle.x = 90, pt.size = 0, base.size = 10, legend.position = "right", 
                plot.title = "scDblFinder.class", colors.use = c("gray50","darkred"), legend.key.size = 0.3)
p
save_plot(p, filename = file.path(dir.results, "figures/violin_seurat_scDblFinder"),  width = 11, height = 4, formats = c("pdf","png"))

```

### Remove doublets

```{r fig.width=5, fig.height=4}

Idents(obj) <- "scDblFinder.class"
obj <- subset(obj, idents = "FALSE")
Idents(obj) <- "seurat_clusters"
obj <- subset(obj, idents = c(56,57,58), invert = T)

Dim_Plot(object = obj, group.by = "scDblFinder.class", label.show = F, label.size = 5, legend = T, 
         dark.theme = F, figure.plot = T, legend.position = "top", cell.countours = F, colors.use = c("gray50","darkred"))

```


```{r fig.width=10, fig.height=4, warning=F, message=F}

Stacked_VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mito"), group.by = "seurat_clusters", 
                angle.x = 90, pt.size = 0.1, base.size = 10, legend.position = "none")

```


### Clean cluster 20

```{r fig.width=4, fig.height=4}

Idents(obj) <- "seurat_clusters"
C20 <- subset(obj, idents = c(20))
Dim_Plot(object = C20, group.by = "seurat_clusters")
C20 <- subset(C20, subset = nFeature_RNA > 1000 & nFeature_RNA < 3000 & nCount_RNA > 1000 & nCount_RNA < 5000 & percent.mito < 1)
Stacked_VlnPlot(C20, features = c("nFeature_RNA", "nCount_RNA", "percent.mito"), group.by = "seurat_clusters", 
                angle.x = 90, pt.size = 0.1, base.size = 10, legend.position = "none")

```


```{r fig.width=8, fig.height=4}

#sub <- Run_Quality_Control(seurat_input = obj, min_features = 1000, min_counts = 1000, filter_method = "MAD", MAD = 5)
sub <- subset(obj, idents = 20, invert = T)
Dim_Plot(object = sub, group.by = "seurat_clusters", label.show = T, figure.plot = T, legend = F)
obj <- subset(obj, cells = c(WhichCells(sub), WhichCells(C20)))
rm(sub, C20)
p <- Stacked_VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mito"), group.by = "seurat_clusters", 
                angle.x = 90, pt.size = 0.1, base.size = 10, legend.position = "none")
p
save_plot(p, filename = file.path(dir.results, "figures/violin_cluster20_filt"),  width = 8, height = 4, formats = c("pdf","png"))

```


### Highlight_Each_Cluster

```{r fig.width=8, fig.height=6}

p <- Dim_Plot(object = obj, group.by = "seurat_clusters", label.show = T, legend = F, theme.bw = F, figure.plot = T, label.size = 4, pt.size = 0.5) 
p
save_plot(p, filename = file.path(dir.results, "figures/umap_seurat"),  width = 8, height = 6, formats = c("pdf","png"))

```


```{r fig.width=14, fig.height=18}

invisible(gc())

p <- Highlight_Each_Cluster(
  object = obj,
  group.by = "seurat_clusters", 
  highlight.size = 0.5,
  ncol = 5,  
  fontsize = 14,
  custom.colors = NULL
)

p
save_plot(p, filename = file.path(dir.results, "figures/Highlight_Each_seurat_clusters"),  width = 14, height = 18, formats = c("pdf","png"))

```



# Add demuxafy results

```{r fig.width=8, fig.height=6, warning=F, message=F}

# From vireo donor match
#FN_S1256: S1=d0, S2=d2, S5=d3, S6=d1
#FN_3478: S3=d1, S4=d3, S7=d2,S8=d0

# Souporcell
# FN_S1256: 
# 0 = S2 = donor1, 
# 1 = S6 = donor3, 
# 2 = S1 = donor2, 
# 3 = S5 = donor0
# FN_S3478: 
# 0 = S3 = donor2, 
# 1 = S7 = donor0, 
# 2 = S8 = donor1, 
# 3 = S4 = donor3

obj@meta.data[['souporcell']] <- paste0(obj@meta.data$Souporcell_Individual_Assignment,"_", obj@meta.data[,"orig.ident"])  

obj@meta.data$sample <- NULL
obj@meta.data[obj@meta.data$souporcell %in% c("donor0_FN_S1256"),"sample"] <-"S5"
obj@meta.data[obj@meta.data$souporcell %in% c("donor1_FN_S1256"),"sample"] <-"S2"
obj@meta.data[obj@meta.data$souporcell %in% c("donor2_FN_S1256"),"sample"] <-"S1"
obj@meta.data[obj@meta.data$souporcell %in% c("donor3_FN_S1256"),"sample"] <-"S6"

obj@meta.data[obj@meta.data$souporcell %in% c("donor0_FN_S3478"),"sample"] <-"S7"
obj@meta.data[obj@meta.data$souporcell %in% c("donor1_FN_S3478"),"sample"] <-"S8"
obj@meta.data[obj@meta.data$souporcell %in% c("donor2_FN_S3478"),"sample"] <-"S3"
obj@meta.data[obj@meta.data$souporcell %in% c("donor3_FN_S3478"),"sample"] <-"S4"


obj[["bmi_donor"]]=""
obj@meta.data[obj@meta.data$sample %in% c("S1"),"bmi_donor"] <-"S1-PCW7" 
obj@meta.data[obj@meta.data$sample %in% c("S2"),"bmi_donor"] <-"S2-PCW8"
obj@meta.data[obj@meta.data$sample %in% c("S3"),"bmi_donor"] <-"S3-PCW10"
obj@meta.data[obj@meta.data$sample %in% c("S5"),"bmi_donor"] <-"S5-PCW10"
obj@meta.data[obj@meta.data$sample %in% c("S4"),"bmi_donor"] <-"S4-OB-PCW10.5"
obj@meta.data[obj@meta.data$sample %in% c("S6"),"bmi_donor"] <-"S6-OB-PCW12"
obj@meta.data[obj@meta.data$sample %in% c("S7"),"bmi_donor"] <-"S7-PCW12"
obj@meta.data[obj@meta.data$sample %in% c("S8"),"bmi_donor"] <-"S8-PCW12"

obj[["donor"]]=""
obj@meta.data[obj@meta.data$sample %in% c("S1"),"donor"] <-"S1-PCW7" 
obj@meta.data[obj@meta.data$sample %in% c("S2"),"donor"] <-"S2-PCW8"
obj@meta.data[obj@meta.data$sample %in% c("S3"),"donor"] <-"S3-PCW10"
obj@meta.data[obj@meta.data$sample %in% c("S5"),"donor"] <-"S5-PCW10"
obj@meta.data[obj@meta.data$sample %in% c("S4"),"donor"] <-"S4-PCW10.5"
obj@meta.data[obj@meta.data$sample %in% c("S6"),"donor"] <-"S6-PCW12"
obj@meta.data[obj@meta.data$sample %in% c("S7"),"donor"] <-"S7-PCW12"
obj@meta.data[obj@meta.data$sample %in% c("S8"),"donor"] <-"S8-PCW12"


obj[["PCW"]]=""
obj@meta.data[obj@meta.data$sample %in% c("S1"),"PCW"] <-"PCW7" 
obj@meta.data[obj@meta.data$sample %in% c("S2"),"PCW"] <-"PCW8"
obj@meta.data[obj@meta.data$sample %in% c("S3", "S4","S5"),"PCW"] <-"PCW10"
obj@meta.data[obj@meta.data$sample %in% c("S6", "S7","S8"),"PCW"] <-"PCW12"

obj[["age"]]=""
obj@meta.data[obj@meta.data$sample %in% c("S1"),"age"] <-"7" 
obj@meta.data[obj@meta.data$sample %in% c("S2"),"age"] <-"8"
obj@meta.data[obj@meta.data$sample %in% c("S3", "S4","S5"),"age"] <-"10"
obj@meta.data[obj@meta.data$sample %in% c("S6", "S7","S8"),"age"] <-"12"


obj[["BMI"]]=""
obj@meta.data[obj@meta.data$sample %in% c("S1"),"BMI"] <-"PCW7" 
obj@meta.data[obj@meta.data$sample %in% c("S2"),"BMI"] <-"PCW8"
obj@meta.data[obj@meta.data$sample %in% c("S3","S5"),"BMI"] <-"PCW10"
obj@meta.data[obj@meta.data$sample %in% c("S4"),"BMI"] <-"OB-PCW10.5"
obj@meta.data[obj@meta.data$sample %in% c("S6"),"BMI"] <-"OB-PCW12"
obj@meta.data[obj@meta.data$sample %in% c("S7","S8"),"BMI"] <-"PCW12"

obj[["group"]]=""
obj@meta.data[obj@meta.data$sample %in% c("S1","S2"),"group"] <-"NORMAL-7-8" 
obj@meta.data[obj@meta.data$sample %in% c("S3","S5","S7","S8"),"group"] <-"NORMAL-10-12"
obj@meta.data[obj@meta.data$sample %in% c("S4","S6"),"group"] <-"OBESE-10-12"

obj[["sex"]]=""
obj@meta.data[obj@meta.data$sample %in% c("S2","S5","S7"),"sex"] <- "F" 
obj@meta.data[obj@meta.data$sample %in% c("S1","S3","S4","S6","S8"),"sex"] <- "M"

obj@meta.data$donor_order <- factor(obj@meta.data$donor, levels = c("S1-PCW7","S2-PCW8","S3-PCW10","S5-PCW10","S4-PCW10.5","S6-PCW12", "S7-PCW12","S8-PCW12"))
obj@meta.data$BMI_order <- factor(obj@meta.data$BMI, levels = c("PCW7","PCW8","PCW10","OB-PCW10.5", "PCW12", "OB-PCW12"))
obj@meta.data$PCW_order <- factor(obj@meta.data$PCW, levels = c("PCW7","PCW8","PCW10", "PCW12"))
obj@meta.data$group_order <- factor(obj@meta.data$group, levels = c("NORMAL-7-8","NORMAL-10-12","OBESE-10-12"))

```

# Save post-processed data

```{r}

saveRDS(obj, file.path(dir.results, "rdata/FN_postprocessed.rds"))

```


# Session Info

```{r session_info}

utils::capture.output(devtools::session_info())

```

