---
title: "000-HUDECA — 002-annotation"
output: html_document
date: '`r Sys.Date()`'
---

## Goal
Cell type annotation

## Setup
```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
root <- '/Users/yvon.mbouamboua/projects/singlecell/000-hudeca'
outdir <- file.path(root, 'out', '0002-annotation')
invisible(lapply(c(outdir, file.path(outdir, 'data'), file.path(outdir, 'plots')),
                 dir.create, recursive = TRUE, showWarnings = FALSE))
srcr <- file.path(root, 'src', 'R')
invisible(lapply(list.files(srcr, full.names = TRUE), source))
appdir <- Sys.getenv('APPS', '~/projects/apps')
if (dir.exists(appdir)) {
  pkgfile <- file.path(appdir, 'packages.R')
  if (file.exists(pkgfile)) sys.source(pkgfile, envir = .GlobalEnv)
  helper_files <- list.files(appdir, pattern = '\\.R$', full.names = TRUE)
  helper_files <- helper_files[basename(helper_files) != 'packages.R']
  invisible(lapply(helper_files, sys.source, envir = .GlobalEnv))
}
```

# Load integrated data

```{r}

obj <- readRDS(file.path(dir.main, "out/001-integration/data/FN_integrated.rds"))
scCustomize::as.anndata(x = obj, file_path = file.path(outdir, "data"),file_name = "FN_integrated.h5ad")

# Define metadata table
meta <- data.frame(
  BulkSample = c("S1-PCW7","S2-PCW8","S3-PCW10","S5-PCW10","S4-PCW10.5","S6-PCW12","S7-PCW12","S8-PCW12"),
  sample = c("S1-PCW7","S2-PCW8","S3-PCW10","S5-PCW10","S4-PCW10.5","S6-PCW12", "S7-PCW12","S8-PCW12"),
  stage = c("PCW7","PCW8","PCW10","PCW10","PCW10","PCW12","PCW12","PCW12"),
  group = c("PCW7-8","PCW7-8","PCW10","PCW10","PCW10","PCW12","PCW12","PCW12"),
  age = c("7","8","10","10","10","12","12","12"),
  bmi = c("Normal","Normal","Normal","Obese","Normal","Obese","Normal","Normal"),
  bmi_group = c("Normal-7-8","Normal-7-8","Normal-10-12","Obese-10-12","Normal-10-12","Obese-10-12","Normal-10-12","Normal-10-12"),
  sex = c("Male","Female","Male","Male","Female","Male","Female","Male"),
  stringsAsFactors = FALSE
)

# Convert to factors with specified levels
meta$sample <- factor(meta$sample, levels = c("S1-PCW7","S2-PCW8","S3-PCW10","S5-PCW10","S4-PCW10.5","S6-PCW12","S7-PCW12","S8-PCW12"))
meta$stage <- factor(meta$stage, levels = c("PCW7","PCW8","PCW10","PCW12"))
meta$group <- factor(meta$group, levels = c("PCW7-8","PCW10","PCW12"))
meta$age <- factor(meta$age, levels = c("7","8","10","12"))
meta$bmi <- factor(meta$bmi, levels = c("Normal","Obese"))
meta$bmi_group <- factor(meta$bmi_group, levels = c("Normal-7-8","Normal-10-12","Obese-10-12"))
meta$sex <- factor(meta$sex, levels = c("Male","Female"))

# Add metadata to Seurat object
obj <- AddMetaData(obj, meta = meta[match(obj$BulkSample, meta$BulkSample), ])

cellmap(object = obj, group.by = "sample")
cellmap(object = obj, group.by = "sex")
cellmap(object = obj, group.by = "age")
cellmap(object = obj, group.by = "stage")
cellmap(object = obj, group.by = "group")
cellmap(object = obj, group.by = "bmi")
cellmap(object = obj, group.by = "bmi_group")
cellmap(object = obj, group.by = "RNA_snn_res.0.5", label = T)

```


```{r}

df1 <- raw@meta.data %>%
  group_by(BulkSample) %>%
  summarise(n_cells = n()) %>%
  mutate(stage = "raw")


df2 <- obj@meta.data %>%
  group_by(sample) %>%
  summarise(n_cells = n()) %>%
  mutate(stage = "pre-integration")

df1
df2

post_int <- integrated@meta.data %>%
  group_by(sample) %>%
  summarise(n_cells = n()) %>%
  mutate(stage = "post-integration")

# Combine
df <- bind_rows(df1,df2)

```



# Azimuth prediction

```{r fig.width=14, fig.height=8, message=F, warning=F}

invisible(gc())
DefaultAssay(obj) <- "RNA"
obj <- FindVariableFeatures(obj)
obj <- RunAzimuth(obj, reference = "fetusref")
invisible(gc())
DefaultAssay(obj) <- "RNA"
obj@reductions$ref.umap <- NULL
obj[["refAssay"]] <- NULL
obj[["nCount_refAssay"]] <- NULL
obj[["nFeature_refAssay"]] <- NULL
obj[["percent.mt"]] <- NULL

#Dim_Plot(object = obj, group.by = "seurat_clusters", label = T, legend = F, label.size = 4, pt.size = 1) 
#Dim_Plot(object = obj, group.by = "predicted.annotation.l1", label = T, legend = F, label.size = 4, pt.size = 1) 
#Stacked_VlnPlot(object = obj, features = c("nCount_RNA","nFeature_RNA", "percent.mito"),angle.x = 90, pt.size = 0.005,base.size = 12)

p <- Cell_Prediction_Heatmap(obj, 
                        cluster_var = "seurat_clusters", 
                        annotation_var = "predicted.annotation.l1",
                        theme_color = "viridis",
                        cluster_rows = F,
                        cluster_cols = T)
print(p)
save_plot(p, filename = file.path(outdir, "plots/azimuth_prediction"),formats = c("png", "pdf"),  width = 14, height = 10)

```



# Automated cellTypist annotation

```{r fig.width=14, fig.height=4}

# Chenqu_Suo_2022
h5ad_obs <- readr::read_csv(file.path(outdir, "data/FN_Celltypist_Pan_Fetal_Human.csv"))
dim(h5ad_obs)
dim(obj)
h5ad_obs <- h5ad_obs[, c("predicted_labels", "over_clustering", "majority_voting","conf_score")]
names(h5ad_obs) <- c("Chenqu_Suo_2022_predicted_labels", "Chenqu_Suo_2022_over_clustering", "Chenqu_Suo_2022_majority_voting","Chenqu_Suo_2022_conf_score")
metadata <- obj@meta.data
metadata <- cbind(metadata, h5ad_obs)
obj@meta.data <- metadata
dim(obj)

cell_prediction_heatmap(obj, cluster_var = "seurat_clusters", annotation_var = "Chenqu_Suo_2022_majority_voting",
                        theme_color = "viridis",cluster_rows = T,cluster_cols = F, plot_title = "Pan_Fetal_Human")

# Braun_2023
h5ad_obs <- readr::read_csv(file.path(outdir, "data/FN_Celltypist_Developing_Human_Brain.csv"))
dim(h5ad_obs)
dim(obj)
h5ad_obs <- h5ad_obs[, c("predicted_labels", "over_clustering", "majority_voting","conf_score")]
names(h5ad_obs) <- c("Braun_2023_predicted_labels", "Braun_2023_over_clustering", "Braun_2023_majority_voting","Braun_2023_conf_score")
metadata <- obj@meta.data
metadata <- cbind(metadata, h5ad_obs)
obj@meta.data <- metadata
dim(obj)

cell_prediction_heatmap(obj, cluster_var = "seurat_clusters", annotation_var = "Braun_2023_majority_voting",
                        theme_color = "viridis",cluster_rows = T,cluster_cols = F, plot_title = "Developing_Human_Brain")

cell_prediction_heatmap(obj, cluster_var = "seurat_clusters", annotation_var = "Chenqu_Suo_2022_predicted_labels",
                        theme_color = "viridis",cluster_rows = T,cluster_cols = F, plot_title = "Pan_Fetal_Human")

```


# umap

```{r fig.width=8, fig.height=6}

cellmap(obj, group.by = "seurat_clusters", label = T, dark = T, legend = T, pt.size = 0.5)

```



# Check cluster quality

```{r fig.width=10, fig.height=5}
cellvio(obj, group.by = "seurat_clusters", features = c("nFeature_RNA", "nCount_RNA","percent_mito"), stack = T, show.pval = F)
cellvio(obj, group.by = "sample", features = c("nFeature_RNA", "nCount_RNA","percent_mito"), stack = T, pt.size = 0.1)
```

# Find markers

```{r fig.width=10, fig.height=20, warning=FALSE, message=FALSE,error=FALSE}

## Filter quality by removing unwanted genes
sub <- filter_genes(obj)
markers <- cellmarker(sub,  group.by = "seurat_clusters", only.pos = T, max.cells.per.ident = 1000, min.pct = 0.1)
write.table(markers, file.path(outdir, "data/harmony_clusters.tsv"), sep = "\t", row.names = F, quote = F)
rm(sub)
markers <- read.delim(file.path(outdir, "data/harmony_clusters.tsv"))
top <- markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC)
celldot(obj, features = rev(top$gene), flip = T, group.by = "seurat_clusters") 

```


```{r fig.width=10, fig.height=6}

Idents(obj) <- "seurat_clusters"
obj <- FindSubCluster(obj, cluster = 11, resolution = 0.3, graph.name = "RNA_snn", subcluster.name = "C11")
Idents(obj) <- "C11"
obj <- FindSubCluster(obj, cluster = 23, resolution = 0.2, graph.name = "RNA_snn", subcluster.name = "subclusters")
cellmap(object = obj, group.by = "subclusters", label = T, pt.size = 1, n.cells = T, label.size = 5)

```



```{r fig.width=12, fig.height=6}

gnrh1 <- cellpick(obj, clusters = c(27), cluster.col = "seurat_clusters", pos.genes =  c("GNRH1"), expr.thresh = 0)

marker_panel <- list(
  OEC = c("SOX10","S100B","PLP1","PMP22","CLDN11","GFAP"),
  Schwann = c("MPZ","MBP","ERBB3","SOX10","PLP1","GINS3","IL1RAP"),
  Melanocyte = c("MITF","TYR","PMEL","DCT","MLANA")
  #Neuron = c("SLC24A2","KLHL1","PRDM8","OMP","TUBB3","GAP43")
)

marker_panel <- unique(unlist(marker_panel))
celldot(obj, features = c(marker_panel,"PDZRN4","ADAM12"), flip = T, base.size = 9)

mel <- cellpick(obj, clusters = c("11_0","11_3"), cluster.col = "subclusters", pos.genes =  c("MITF"), expr.thresh = 1)
# sch.cells <- cellpick(obj, clusters = c("29_0"), cluster.col = "subclusters", genes =  c("MPZ"), threshold = 1)

```




```{r}

cellmap(object = obj, group.by = "subclusters", label = T, label.face = "bold", label.size = 5, 
         pt.size = 0.5, legend = F, dark = T, figplot = T)
```


# Label transfert

## Annotate OHBC

```{r}

cellvio(obj, group.by = "sample", features = c("nCount_RNA"))
cellmap(obj, group.by = "RNA_snn_res.0.5", label = T)

ohbc <- cellpick(
  obj,
  pos.genes = c("TP63","KRT5","KRT14","COL17A1","DST","MEG3"),
  min.genes = 2,
  expr.thresh = 0.5,
  neg.genes = c("MUC5AC","MUC5B","CDC20B","SOX2"),
  cluster.col = "RNA_snn_res.0.5",
  clusters = 13,
  meta.col = "group",
  meta.vals = c("PCW10","PCW12")
)

```

## Annotation level 2

```{r fig.width=10, fig.height=6}

obj@meta.data$ann2 <- NULL
obj@meta.data[obj@meta.data$seurat_clusters %in% c(28),"ann2"] <- "MSC" # PAX6, "SOX2
obj@meta.data[obj@meta.data$seurat_clusters %in% c(32),"ann2"] <-"Deut" # "DEUP1","HES6","CCNO"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(7,41),"ann2"] <-"MCC" # "SNTN","DNAH5"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(10,35,36),"ann2"] <-"EC"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(34),"ann2"] <-"Lymph"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(19),"ann2"] <-"Peri" # PDGFRB
obj@meta.data[obj@meta.data$seurat_clusters %in% c(22,40),"ann2"] <-"NP" # "GFAP",SOX2,HES5
obj@meta.data[obj@meta.data$seurat_clusters %in% c(27),"ann2"] <-"GABA" #"SLC32A1","GAD1","GAD2"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(21,39),"ann2"] <-"GLUT" # "SLC17A6","SLC17A7","CRH"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(38),"ann2"] <-"NOS1"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(11),"ann2"] <- "OEC" # "SOX10", "S100B", "VIM", "PLP1", "GFAP", "FABP7", "CD44","NELL2"
obj@meta.data[obj@meta.data$subclusters %in% c("11_1"),"ann2"] <-"NCC" 
obj@meta.data[obj@meta.data$Braun_2023_majority_voting %in% c("Head neural crest cells"),"ann2"] <- "NCC" #"FOXD3", "TFAP2A", "NGFR"
obj@meta.data[obj@meta.data$subclusters %in% c("11_4"),"ann2"] <-"SCP" 
obj@meta.data[obj@meta.data$subclusters %in% c("11_2"),"ann2"] <-"SC" # "MPZ","MBP","ERBB3"
obj@meta.data[rownames(obj@meta.data) %in% c(mel),"ann2"] <-"Mel" # "MITF"
obj@meta.data[obj@meta.data$Chenqu_Suo_2022_predicted_labels %in% c("MELANOCYTE"),"ann2"] <- "Mel" 
obj@meta.data[obj@meta.data$seurat_clusters %in% c(9,18,29),"ann2"] <-"Cartl" # "COL9A1", "COL11A1", "COL2A1", "ACAN", "WWP2"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(1,6,14),"ann2"] <-"Osteo" # "RUNX2"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(4,37),"ann2"] <-"Mes0" # prolif cells
obj@meta.data[obj@meta.data$seurat_clusters %in% c(0,5,13),"ann2"] <- "Mes1"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(2,20,24,30,33),"ann2"] <- "Mes2"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(12),"ann2"] <-"SM"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(15),"ann2"] <- "Sat"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(42),"ann2"] <-"GBC" # "HES6","TCF12","TCF4","TCF3","HIF1A","ARNT"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(8,16,31),"ann2"] <-"RHBC" # TP63, SOX2
#obj@meta.data[obj@meta.data$seurat_clusters %in% c(16),"ann2"] <-"0HBC"
obj@meta.data[rownames(obj@meta.data) %in% c(ohbc),"ann2"] <- "OHBC" # TP63, COL17A1
obj@meta.data[obj@meta.data$seurat_clusters %in% c(26),"ann2"] <-"CycBasal"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(3),"ann2"]  <- "MV" # "MUC5AC (PMCID: PMC8564600), "GLIS3
obj@meta.data[obj@meta.data$seurat_clusters %in% c(17),"ann2"]  <- "SUS" # "IQCJ-SCHIP1", "TMEM132D", "LPGAT1"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(25),"ann2"] <-"MG" # "CX3CR1", "CD14","TMEM119","MS4A4A","CSF1R","ITGAM","P2RY12"
obj@meta.data[obj@meta.data$seurat_clusters %in% c(23),"ann2"] <- "iOSN"
obj@meta.data[obj@meta.data$subclusters %in% c("23_1"),"ann2"] <- "INP" # NEUROD1, NEUROG1, LHX2
obj@meta.data[rownames(obj@meta.data) %in% c(gnrh1),"ann2"] <-"GnRH" # "ISL1","GNRH1","LINC01060","TAC3

cellmap(obj, group.by="ann2",  label = T, pt.size = 0.5, figplot = T, dark = T)

```




```{r fig.width=18, fig.height=6}

pb_prop_plot(input = obj, donor.col = "sample", stage.col = "stage", cell.col  = "ann2",
             ncol = 7, show.p = F, leg.pos = "none",
  plot.type = "bar",pt.size = 1.5, facet.by = "cell", font.size = 10, facet.bg = F)

```


# Order the cells

```{r}

# Level 2: fine-grained cell types
ann2_levels <- c(
"RHBC","OHBC","CycBasal","GBC","INP","iOSN","Deut","MCC","MV","SUS",  # Epithelium
"NP","GLUT","GABA","GnRH","NOS1", # Neuronal
"NCC","SCP","SC","OEC","Mel", # Neural Crest / Glia
"MSC","Mes0","Mes1","Mes2","Osteo","Cartl", # Mesenchyme
"Peri","EC","Lymph",  # Vascular
"Sat","SM", # Muscle
"MG" # Immune
)

# Ensure your Seurat object has ann2
obj$ann2 <- factor(obj$ann2, levels = ann2_levels)

# Level 1 mapping → lineage groups

ann_map_level1 <- list(
  "Olfactory Epithelium"   = c("OHBC","CycBasal","GBC","MV","SUS","INP","iOSN"),
  "Respiratory Epithelium" = c("RHBC","Deut","MCC"),
  "Neuronal"               = c("NP","GLUT","GABA","GnRH","NOS1"),
  "NeuralCrest_Glia"       = c("NCC","SCP","SC","OEC","Mel"),
  "Mesenchyme"             = c("MSC","Mes0","Mes1","Mes2","Osteo","Cartl"),
  "Vascular"               = c("Peri","EC","Lymph"),
  "Muscle"                 = c("Sat","SM"),
  "Immune"                 = c("MG")
)

# Invert mapping (cell type → lineage)
inv_level1 <- unlist(lapply(names(ann_map_level1), function(x) {
  setNames(rep(x, length(ann_map_level1[[x]])), ann_map_level1[[x]])
}))

level1_vector <- inv_level1[as.character(obj$ann2)]
names(level1_vector) <- colnames(obj)

# Keep only cells with valid mapping
valid_cells1 <- names(level1_vector)[!is.na(level1_vector)]
#obj <- subset(obj, cells = valid_cells1)

obj <- AddMetaData(obj, metadata = level1_vector[valid_cells1], col.name = "ann1")
obj$ann1 <- factor(obj$ann1, levels = names(ann_map_level1))

ann_map_level0 <- list(
  "Epithelium" = c("Olfactory Epithelium","Respiratory Epithelium"),
  "Neuronal"   = c("Neuronal"),
  "NeuralCrest_Glia" = c("NeuralCrest_Glia"),
  "Mesenchyme" = c("Mesenchyme"),
  "Vascular"   = c("Vascular"),
  "Muscle"     = c("Muscle"),
  "Immune"     = c("Immune")
)

# inv_level0 maps Level1 → Level0
inv_level0 <- unlist(lapply(names(ann_map_level0), function(x) {
  setNames(rep(x, length(ann_map_level0[[x]])), ann_map_level0[[x]])
}))

# Create a vector for ALL cells
level0_vector <- inv_level0[as.character(obj$ann1)]

# Name the vector with Seurat cell names
names(level0_vector) <- colnames(obj)

# Add to Seurat metadata
obj <- AddMetaData(obj, metadata = level0_vector, col.name = "ann0")

# Convert to factor with only defined levels
obj$ann0 <- factor(obj$ann0, levels = names(ann_map_level0))

#saveRDS(obj, file.path(dir.main, "output/002-annotation/data/FN_harmony.rds"))

```

```{r}
# # Level 2: fine-grained cell types
ann2_levels <- c(
"RHBC","OHBC","CycBasal","GBC","INP","iOSN","Deut","MCC","MV","SUS",  # Epithelium
"NP","GLUT","GABA","GnRH","NOS1", # Neuronal
"NCC","SCP","SC","OEC","Mel", # Neural Crest / Glia
"MSC","Mes0","Mes1","Mes2","Osteo","Cartl", # Mesenchyme
"Peri","EC","Lymph",  # Vascular
"Sat","SM", # Muscle
"MG" # Immune
)

# Ensure your Seurat object has ann2
obj$ann2 <- factor(obj$ann2, levels = ann2_levels)

# Level 1 mapping → lineage groups

ann_map_level1 <- list(
  "Olfactory Epithelium"   = c("OHBC","CycBasal","GBC","MV","SUS","INP","iOSN"),
  "Respiratory Epithelium" = c("RHBC","Deut","MCC"),
  "Neuronal"               = c("NP","GLUT","GABA","GnRH","NOS1"),
  "NCC-Glia"               = c("NCC","SCP","SC","OEC","Mel"),
  "Mesenchyme"             = c("MSC","Mes0","Mes1","Mes2","Cartl"),
  "Osteoblast"             = c("Osteo"),
  "Vascular"               = c("Peri","EC","Lymph"),
  "Muscle"                 = c("Sat","SM"),
  "Immune"                 = c("MG")
)

# Invert mapping (cell type → lineage)
inv_level1 <- unlist(lapply(names(ann_map_level1), function(x) {
  setNames(rep(x, length(ann_map_level1[[x]])), ann_map_level1[[x]])
}))

level1_vector <- inv_level1[as.character(obj$ann2)]
names(level1_vector) <- colnames(obj)

# # Keep only cells with valid mapping
valid_cells1 <- names(level1_vector)[!is.na(level1_vector)]
#obj <- subset(obj, cells = valid_cells1)

obj <- AddMetaData(obj, metadata = level1_vector[valid_cells1], col.name = "ann1")
obj$ann1 <- factor(obj$ann1, levels = names(ann_map_level1))

ann_map_level0 <- list(
  "Epithelium" = c("Olfactory Epithelium","Respiratory Epithelium"),
  "Neuronal"   = c("Neuronal"),
  "NCC-Glia" = c("NCC-Glia"),
  "Mesenchyme" = c("Mesenchyme","Osteoblast"),
  "Vascular"   = c("Vascular"),
  "Muscle"     = c("Muscle"),
  "Immune"     = c("Immune")
)

# inv_level0 maps Level1 → Level0
inv_level0 <- unlist(lapply(names(ann_map_level0), function(x) {
  setNames(rep(x, length(ann_map_level0[[x]])), ann_map_level0[[x]])
}))

# Create a vector for ALL cells
level0_vector <- inv_level0[as.character(obj$ann1)]

# Name the vector with Seurat cell names
names(level0_vector) <- colnames(obj)

# Add to Seurat metadata
obj <- AddMetaData(obj, metadata = level0_vector, col.name = "ann0")

# Convert to factor with only defined levels
obj$ann0 <- factor(obj$ann0, levels = names(ann_map_level0))

saveRDS(obj, file.path(dir.main, "output/002-annotation/data/FN_harmony_clean.rds"))
# scCustomize::as.anndata(x = obj, file_path = file.path(dir.main, "output/002-annotation/data"),file_name = "hudeca_nose.h5ad")

cellmap(obj, group.by = "ann0")
cellmap(obj, group.by = "ann1") 
cellmap(obj, group.by = "ann2")
```




# Remove outlier cells

```{r fig.width=12, fig.height=14}

highlight_clusters(object = obj, group.by = "ann2", mode = "single", ncol = 4)

res <- filter_outliers(obj = obj, group.by = "ann2", density.threshold = 0.95, k = 40, relax = 3,
                       clean.all = T, verbose = T, return.plot = F)

highlight_clusters(object = res$obj, group.by = "ann2", mode = "single", ncol = 4)

obj <- res$obj

res <- filter_outliers(obj = obj, group.by = "ann2", target.clusters = "MSC",
                       density.threshold = 0.70, k = 50, relax = 1.2,clean.all = F, verbose = T, return.plot = F)
obj <- res$obj
rm(res)
highlight_clusters(object = obj, group.by = "ann2",cluster.names = "MSC", mode = "single", ncol = 1)


res <- filter_outliers(obj = obj, group.by = "ann2", target.clusters = "Peri",
                       density.threshold = 0.90, k = 50, relax = 2,clean.all = F, verbose = T, return.plot = F)

highlight_clusters(object = res$obj, group.by = "ann2",cluster.names = "Peri", mode = "single", ncol = 1)

obj <- res$obj
rm(res)
highlight_clusters(object = obj, group.by = "ann2", mode = "single", ncol = 4)

```



```{r fig.width=10, fig.height=6}

cellmap(obj, group.by="ann0",  label = T, pt.size = 0.5, figplot = T, dark = T)
cellmap(obj, group.by="ann1",  label = T, pt.size = 0.5, figplot = T, dark = T)
cellmap(obj, group.by="ann2",  label = T, pt.size = 0.5, figplot = T, dark = T)
cellmap(obj, group.by="sample",  label = F, pt.size = 0.5, figplot = T, dark = T)
cellmap(obj, group.by="group",  label = T, pt.size = 0.5, figplot = T, dark = T)
cellmap(obj, group.by="bmi",  label = T, pt.size = 0.5, figplot = T, dark = T)

```


# Save data

```{r}

obj <- fix_seurat_matrix_names(obj)
saveRDS(obj, file.path(outdir, "data/FN_harmony.rds"))

```


# Remove ambient RNA

```{r}

decontx <- run_decontx(obj)
saveRDS(decontx, file.path(outdir, "data/decontX.rds"))
obj[["RNA"]]@layers$counts <- decontx$cleaned_counts
obj$decontX_contamination <- decontx$sce$decontX_contamination
obj <- NormalizeData(obj) %>% FindVariableFeatures() %>% ScaleData()
obj <- fix_seurat_matrix_names(obj)

```

## Plot decontx contamination

```{r}

df <- data.frame(
  UMAP1 = obj[["umap"]]@cell.embeddings[,1],
  UMAP2 = obj[["umap"]]@cell.embeddings[,2],
  contamination = decontx$sce$decontX_contamination
)

p <- ggplot(df, aes(UMAP1, UMAP2, color = contamination)) +
  geom_point(size = 0.3, alpha = 0.8) +
  #scale_color_viridis_c() +
  scale_color_gradientn(
    colors = c("blue", "green", "yellow", "orange", "red"),
    guide = guide_colorbar(title = "Contamination", title.position = "left",
      title.theme = element_text(angle = 90, vjust = 0.5),
      barwidth = 1, barheight = 8, ticks = TRUE, frame.colour="black", ticks.colour="black")) +
  plot_theme(theme = "classic", font.size = 14) +
  ggtitle("DecontX contamination") +
  theme(legend.pos = "right", legend.title.align = 0.5)
p
save_plot(p, file.path(outdir, "plots/cecontx_contamination"), formats = "pdf", width = 6, height = 5)

```


## plot_decontx_marker_percen

```{r fig.width=18, fig.height=6}

markers <- c("XIST","UTY","HBB","HBA1")

res <- plot_decontx_marker_percent(
  sce = decontx$sce,
  markers = markers,fontsize = 14,
  group_by = "ann2",
  assays = c("counts","decontXcounts"),
  plot_type = "bar", custom_colors = c("#151A1F", "#8CB9E6")
)

res$plot
save_plot(res$plot, file.path(outdir, "plots/decontx_marker_percent"), formats = "pdf", width = 18, height = 6)


```



# Sex-chromosome–based sample identity and contamination quality control

```{r}

# Run QC and generate violin plots
res <- sex_contamination_qc(
  obj,
  sample_col = "sample",
  y_genes = c("UTY","RPS4Y1","ZFY","DDX3Y","KDM5D"),
  xist_gene = "XIST",
  eps = 1e-6,
  female_thresh = 1,
  male_thresh = -1,
  eryth_genes = c("HBB","HBA1","HBA2","HBE1","HBG1","HBG2","HBM"),
  eryth_percentile = 0.95,
  remove_ambiguous = T,
  majority_threshold = 0.5,
  min_sex_score_diff = NULL,
  out_file = file.path(outdir, "data/sex_qc.tsv")
)

# Access results
tbl_summary <- res$tbl_summary
obj <- res$filtered
obj <- NormalizeData(obj) %>% FindVariableFeatures() %>% ScaleData()
saveRDS(res$object, file.path(dir.main, "out/002-annotation/data/FN_harmony.rds"))
saveRDS(res$filtered, file.path(dir.main, "out/002-annotation/data/FN_harmony_clean.rds"))
rm(res)
tbl_summary

```


## Session info
```{r}
sessionInfo()
```
