---
title: "10x Hudeca Nasal: Figures"
author: "Mbouamboua Yvon <yvon.mbouamboua@inserm.fr>"
output:
    html_document:
       toc: true
       toc_float: true
       number_sections: true
date: 'Last update: `r format(Sys.time())`'
---

# Setup

```{r setup, include=TRUE, message=FALSE, warning=FALSE}

invisible(gc())
knitr::opts_chunk$set(warning = FALSE, results = TRUE, message = FALSE, comment = "")
dir.main <- "/Users/yvon.mbouamboua/Documents/projects/singlecell/hudeca_nasal"
dir.results <- file.path(dir.main, "results", "010_figures")
dir.create(dir.results, showWarnings = FALSE, recursive = TRUE)

# Load R packages
packages <- c("Seurat", "ggplot2", "dplyr", "tidyverse","grid", "ggpubr","decoupleR","dorothea", "viper","ggraph",
              "SingleCellExperiment", "scran", "scater", "pheatmap", "ComplexHeatmap", "ComplexUpset", "patchwork")

# Function to check and install packages from CRAN or Bioconductor
install_and_load <- function(pkg) {
  # Check if the package is from Bioconductor
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
    BiocManager::install(pkg, ask = FALSE)
  }
  # Load the package
  suppressWarnings(suppressMessages(library(pkg, character.only = TRUE)))
}

# Apply function to each package
suppressPackageStartupMessages({
  invisible(lapply(packages, install_and_load))
})

# Load helper functions
# Define the directory where utility scripts are located
utils_dir <- "~/Documents/projects/singlecell/utilities/"
# Get a list of all R files in the directory
r_files <- list.files(utils_dir, pattern = "\\.R$", full.names = TRUE)
# Source each R file
invisible(suppressWarnings(suppressMessages(lapply(r_files, source))))

```

# Load data

```{r}

obj <- readRDS(file.path(dir.main, "results/003_celltype/rdata/FN.rds"))

```

# Define colors

```{r}

custom_colors <- list()
custom_colors$ann_level_2 <- c("Cartilages" = "#0B4B19","Stromal0" = "#99D6A9", "Stromal1" = "#1B8F76","Stromal2" = "#9DAF07","Osteoblasts" = "#4CAD4C",  "Progenitor cells" = "blue", 
"Lymphatic EC" = "#F78896", "Vascular EC" = "#E788C2","Pericytes" = "#BBD870","Satellites" = "#CB7647","Skeletal muscle" = "#926B54","Neural crest" = "#E3D9AC","Olf. ensh. glia" = "palevioletred3","Schwann precursors" = "orangered","Schwann cells" = "#95ccff","Melanocytes" = "#eb10fd","NOS1 neurons" = "#95819F","Olf. HBCs" = "#E41A1C","Basal KRT" = "#C82C73", 
"Cycling basal KRT" = "#C2A523", "Microvillars" = "#efe13c","Multiciliated" = "#1f618d","Deuterosomal" = "#3498db","Sustentaculars" = "#C09ACA", "GBCs" = "#F48B5A", "INP" = "#E69F00", "iOSNs" = "#f05b43","mOSNs" = "#33b8ff","Neural progenitors" = "#6A0B78","Excitatory neurons" = "#706fd3","Inhibitory neurons" = "#800EF1","GnRH neurons" = "#2EECDB",
"Myeloid" = "#736376","Microglia" = "#91BFB7")
 custom_colors$sample_donor = c("S1-PCW7"="#E6AB02","S2-PCW8"="#A6761D","S3-PCW10"="#7570B3","S4-OB-PCW10.5"="#E7298A",  
 "S5-PCW10"="#66A61E", "S6-OB-PCW12"="#D95F02","S7-PCW12"="#16658C", "S8-PCW12"="#134F0D")
 custom_colors$donor = c("S1-PCW7"="#E6AB02","S2-PCW8"="#A6761D","S3-PCW10"="#7570B3","S4-PCW10.5"="#E7298A",  "S5-PCW10"="#16658C","S6-PCW12"="#D95F02", "S7-PCW12"="#66A61E","S8-PCW12"="#134F0D")
 custom_colors$PCW = c("PCW7"="#E6AB02","PCW8"="#A6761D","PCW10"="#7570B3","PCW12"="#134F0D")
 custom_colors$BMI = c("PCW7"="#E6AB02","PCW8"="#A6761D","PCW10"="#7570B3","OB-PCW10.5"="#E7298A","OB-PCW12"="#D95F02",
 "PCW12"="#134F0D")
 custom_colors$ann_level_1 <- c("Progenitor cells" = "blue","Respiratory epithelium" = "#5562B7", "Olfactory epithelium" = "#EF1B4F","Neurons" = "#6E5489","Glial" = "#919976", "Stroma" = "#009E73","Vasculars" = "#CC79A7","Myocytes" = "#803800",
 "Immune" = "#736376","Pericytes" = "#BBD870")
 custom_colors$group <- c("NORMAL-7-8"="#269BCF","NORMAL-10-12"="#8db600","OBESE-10-12"="#DD5D74")
 custom_colors$sex <- c("M"="#854442", "F"="#be9b7b")
 custom_colors$phase <- c("G1"="#9CD4FA", "G2M"="#f1c232", "S"="#90301B","PostM"="#6aa84f", "Non-Cycling" = "gray10")
 
```

# Figure 1
## Fig 1b

```{r }

p <- DimPlot(object = obj, group.by = "donor_order", split.by = "donor_order", label = F, cols = custom_colors$donor, pt.size = 0.5) 
save_plot(p, filename = file.path(dir.results, "figures/Fig_1b"),formats = c("pdf"),  width = 12, height = 5) 

```

## Fig 1c

```{r }

p <- DimPlot(object = obj, group.by = "donor_order", labels = F, cols = custom_colors$donor)
save_plot(p, filename = file.path(dir.results, "figures/Fig_1c"),  width = 6.5, height = 5, formats = c("pdf"))

```

## Fig 1d

```{r }

p <- DimPlot(object = obj, group.by = "ann_level_2_order", labels = T) 
save_plot(p, filename = file.path(dir.results, "figures/Fig_1d"),  width = 9, height = 6, formats = c("pdf"))

```
## Fig 1e-f

```{r }

p1 <- Summarize_Seurat(object = obj, group.by = "ann_level_2_order", condition.by = "sex", plot.variable = "total_transcripts", legend = T, 
                 legend.ncol = 2, plot.title = "", legend.title = "Gender", reorder.bars = F, remove.x.labels = T,remove.x.title = T,
                 remove.y.title = F,legend.position = c(0.05, 0.90), x.title = "celltype", y.title = "Counts", custom.colors = custom_colors$sex)

p2 <- Stacked_VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mito"), group.by = "ann_level_2_order", x_angle = 45, x_vjust = 1,x_hjust = 1, pt.size = 0, font_size = 12, plot.title = "", colors.use = custom_colors$ann_level_2, line.size = 0.5)

p1$plot+p2+plot_layout(ncol = 1)
save_plot(p1$plot+p2+plot_layout(ncol = 1), filename = file.path(dir.results, "figures/Fig_1e-f"),  width = 10, height = 7, formats = c("pdf"))

```

```{r }

# Epithelium
DefaultAssay(obj) <- "RNA"
Idents(obj) <- "ann_level_0"
sub <- subset(obj, idents = c("Epithelium"))
markers <- FindAllMarkers(sub,  group.by = "ann_level_2", only.pos = T, max.cells.per.ident = 10000)
write.table(markers, file.path(dir.results, "tables/OE_RE_ann_level_2_markers.tsv"), sep = "\t", row.names = F, quote = F)
top <- markers %>% group_by(cluster) %>% top_n(n = 5, wt = avg_log2FC)
Idents(sub) <- "ann_level_2"
p <- Dot_Plot(sub, features = rev(c(top$gene)), flip.axes = T, dot.scale = 6, base.size = 10,plot.title = "Olfactory + Respiratory epithelium", theme.bw = T, col.min = 0, angle.x = 45, legend.position = "right", legend.text.size = 8) 
save_plot(p, filename = file.path(dir.results, "figures/Fig_1g_epithelium"),formats = c("pdf"),  width = 4, height = 8)    

# Neurons, Glial, Neural crest
sub <- subset(obj, idents = c("Neurons","Glial","Neural crest"))
markers <- FindAllMarkers(sub,  group.by = "ann_level_2", only.pos = T, max.cells.per.ident = 10000)
write.table(markers, file.path(dir.results, "tables/OE_RE_ann_level_2_markers.tsv"), sep = "\t", row.names = F, quote = F)
top <- markers %>% group_by(cluster) %>% top_n(n = 5, wt = avg_log2FC)
Idents(sub) <- "ann_level_2"
p <- Dot_Plot(sub, features = rev(c(top$gene)), flip.axes = T, dot.scale = 6, base.size = 10,plot.title = "Glial + Neural crest + Neurons", theme.bw = T, col.min = 0, angle.x = 45, legend.position = "right", legend.text.size = 8) 
save_plot(p, filename = file.path(dir.results, "figures/Fig_1g_neurons_glial_neural_crest"),formats = c("pdf"),  width = 4, height = 8)    

# Mesenchyme, Progenitor cells
sub <- subset(obj, idents = c("Mesenchyme","Progenitor cells"))
markers <- FindAllMarkers(sub,  group.by = "ann_level_2", only.pos = T, max.cells.per.ident = 10000)
write.table(markers, file.path(dir.results, "tables/Mesenchyme_Progenitor_markers.tsv"), sep = "\t", row.names = F, quote = F)
top <- markers %>% group_by(cluster) %>% top_n(n = 5, wt = avg_log2FC)
Idents(sub) <- "ann_level_2"
p <- Dot_Plot(sub, features = rev(c(top$gene)), flip.axes = T, dot.scale = 6, base.size = 10,plot.title = "Progenitor cells + Mesenchyme", 
              theme.bw = T, col.min = 0, angle.x = 45, legend.position = "right", legend.text.size = 8) 
save_plot(p, filename = file.path(dir.results, "figures/Fig_1g_progenitor_mesenchyme"),formats = c("pdf"),  width = 4, height = 8)    

# Vasculars, Myocytes, Immune
sub <- subset(obj, idents = c("Endothelium","Myocytes","Immune"))
markers <- FindAllMarkers(sub,  group.by = "ann_level_2", only.pos = T, max.cells.per.ident = 10000)
write.table(markers, file.path(dir.results, "tables/Endo_Myocytes_Immune_ann_level_2_markers.tsv"), sep = "\t", row.names = F, quote = F)
top <- markers %>% group_by(cluster) %>% top_n(n = 5, wt = avg_log2FC)
Idents(sub) <- "ann_level_2"
p <- Dot_Plot(sub, features = rev(c(top$gene)), flip.axes = T, dot.scale = 6, base.size = 10,plot.title = "Vasculars + Myocytes + Immune", 
              theme.bw = T, col.min = 0, angle.x = 45, legend.position = "right", legend.text.size = 8) 
save_plot(p, filename = file.path(dir.results, "figures/Fig_1g_endo_myocytes_immune"),formats = c("pdf"),  width = 4, height = 8)    

```

# Figure 2
## Fig 2b

```{r}

Idents(obj) <- "ann_level_2"
sub <- Seurat_Subset(obj, cluster_id = c("Olf. HBCs", "Microvillars","Sustentaculars","GBCs","INPs","iOSNs","mOSNs"), 
                     ndims = 5, min.dist = 1)

p <- DimPlot(sub, group.by = "ann_level_2", labels = T, cols = custom_colors$ann_level_2)
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_2b"),  width = 6.5, height = 5, formats = c("pdf"))


```

## Fig 2c

```{r fig.width=6, fig.height=4, warning=FALSE, message=F}

Idents(obj) <- "ann_level_2"
sub <- subset(obj, idents = c("Olf. HBCs", "Microvillars","Sustentaculars","GBCs","INPs","iOSNs","mOSNs"))

p <- plot_cluster_percent(data = sub, label_size = 3, cluster_col = "ann_level_2_order", group_col = "PCW_order",sample_col = "sample", group_colors = custom_colors$PCW,sample_colors = pseudo_rep_colors, legend = FALSE, 
                     multi_p = F,ncol = 4,x_angle = 45, fontsize = 8, y_title = "Proportions (%)", theme_type = "minimal")
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_2c"),  width = 6, height = 4, formats = c("pdf"))

```

## Fig 2d

```{r fig.width=2.8, fig.height=10}

Idents(obj) <- "ann_level_2"
sub <- subset(obj, idents = c("Olf. HBCs", "Microvillars","Sustentaculars","GBCs","INPs","iOSNs","mOSNs"))
progression_markers <- FindAllMarkers(sub, group.by = "PCW", max.cells.per.ident = 10000,test.use = "LR") 
write.table(progression_markers, file.path(dir.main, "results/005_differential_expression/tables/OE_PCW_progression_markers.tsv"), 
            sep = "\t", row.names = F, quote = F)
markers <- read.delim(file.path(dir.main, "results/005_differential_expression/tables/OE_PCW_progression_markers.tsv"))
top <- markers %>% group_by(cluster) %>% top_n(n = 20, wt = avg_log2FC)

genes <- intersect(top$gene, rownames(sub@assays$RNA@counts))
filt <- subset(sub, features = genes)
#sub <- sub[rowSums(sub@assays$RNA@counts > 1) >= 2, ]
avg <- AverageExpression(filt, group.by = "PCW_order", features = genes)
avg <- as.matrix(avg$RNA)
avg <- avg[!(rowSums(is.na(avg))),]
cal_z_score <- function(x){(x-mean(x))/sd(x)}
avg <- t(apply(avg, 1, cal_z_score))
sum(is.na(avg)) 
avg[is.na(avg)] <- 0
avg <- as.matrix(avg)
gene_order <- order(apply(avg, 1, which.max))  # Trie selon la colonne où l'expression est maximale
avg <- avg[gene_order, ] 
dim(avg)
newnames <- lapply(rownames(avg), function(x) bquote(italic(.(x))))

ht <- ComplexHeatmap::Heatmap(avg, 
              row_labels = as.expression(newnames),
              show_column_names = TRUE, 
              show_row_names = TRUE, 
              cluster_rows = FALSE,  # Désactiver le clustering des lignes car déjà ordonné
              cluster_columns = FALSE,
              column_names_centered = TRUE,
              show_row_dend = FALSE,
              column_dend_reorder = FALSE,
              column_names_rot = 0,
              col = viridis::viridis(256),
              border = "black",
              row_names_gp = gpar(fontsize = 8),  
              column_names_gp = gpar(fontsize = 8, fontface = "bold"),  
              column_names_side = "top",
              #rect_gp = gpar(col = "white", lwd = 0.1),
              heatmap_legend_param = list(
                title = "z-score",
                legend_height = unit(0.01, 'cm'),
                legend_width = unit(4.5, 'cm'),
                legend_direction = "horizontal", 
                border = 'black'))

pdf(file.path(dir.results, "figures/Fig_2d.pdf"), useDingbats = F,  width = 2.8, height = 10) 
draw(ht, heatmap_legend_side="bottom", annotation_legend_side="bottom")
dev.off()

```

# Figure 3

## Fig 3a

```{r}

highlight_color = c("GBCs"="#F48B5A","INPs"= "#E69F00","iOSNs"="#f05b43", "mOSNs"="#33b8ff")
Idents(obj) <- "ann_level_2"
p <- Highlighted_Clusters(object = obj, cluster.name = c("GBCs","INPs","iOSNs","mOSNs"),group.by = "ann_level_2", highlight.color = highlight_color, legend = T, plot.title = "") + theme(legend.position = c(0.80, 0.20))
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_3a"),  width = 6.5, height = 5, formats = c("pdf"))

```

## Fig 3b

```{r}

Idents(obj) <- "ann_level_2"
sub <- Seurat_Subset(obj, cluster_id = c("GBCs","INPs","iOSNs","mOSNs"), ndims = 5, min.dist = 1)

p <- Dim_Plot(sub, group.by = "ann_level_2", label.show = T, label.size = 5,legend = F, label.fontface = "bold",
              pt.size = 2, colors.use = custom_colors$ann_level_2, no.axes = T, plot.title = "")
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_3b"),  width = 6.5, height = 5, formats = c("pdf"))

```

## Fig_3c

```{r fig.width=8, fig.height=6}

p <- FeaturePlot(sub, features = c("HES6","SOX2","NNAT","MYBL1", "TOP2A","MKI67", "NEUROG1","NEUROD1","GAP43","GNG8","GNG13","TEX15","GNAL","ADCY3","ACSL6","CNG4","OLIG2","RTP1","CBX8","PTPRR","RTN1"))
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_3c"),  width = 8, height = 6, formats = c("pdf"))

```

## Fig 3d

```{r fig.width=10, fig.height=3, warning=F, message=F}

Idents(obj) <- "ann_level_2"
sub <- Seurat_Subset(obj, cluster_id = c("GBCs","INPs","iOSNs","mOSNs"), ndims = 5, min.dist = 1)

p <- DimPlot(object = sub, group.by = "PCW_order", split.by = "PCW_order", labels = F, cols = custom_colors$PCW) + NoAxes() + NoLegend()
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_3d"),formats = c("pdf"),  width = 10, height = 3) 

```

## Fig 3e

```{r}

Idents(obj) <- "ann_level_2"
sub <- Seurat_Subset(obj, cluster_id = c("Olf. HBCs", "Microvillars","Sustentaculars","GBCs","INP","iOSNs","mOSNs"))

genes <- c("COL27A1","TP63","DSC3", # Olf. HBC
           "IQCJ-SCHIP1","SLCO1A2","SLCO1B1", # Sustentaculars
           "CFTR","MUC5AC", # Microvillars
           "NEUROD1","NEUROG1", "TOP2A","MKI67", 
           "HES6", "EZH2", "COL9A2", # GBCs
           "THSD7B","NKAIN1","KCNH6","UNCX", # INP 
           "TEX15","DPF3","GNG8","GAP43","OLIG2","DCX","SPOCK2","NHLH2","TUBB2", "FSTL5","CALB2","PCSK1N", # iOSNs
           "PTPRR","GNGB1","ACSL6","GNG13","RTP1","MGAT4D" # mOSNs
           )

genes <- intersect(genes, rownames(sub@assays$RNA@counts))
filt <- subset(sub, features = genes)
avg <- AverageExpression(filt, group.by = "ann_level_2_order", features = genes)
avg <- as.matrix(avg$RNA)
avg <- avg[!(rowSums(is.na(avg))),]
cal_z_score <- function(x){(x-mean(x))/sd(x)}
avg <- t(apply(avg, 1, cal_z_score))
sum(is.na(avg)) 
avg[is.na(avg)] <- 0
avg <- as.matrix(avg)
gene_order <- order(apply(avg, 1, which.max))  # Trie selon la colonne où l'expression est maximale
avg <- avg[gene_order, ] 
dim(avg)
newnames <- lapply(rownames(avg), function(x) bquote(italic(.(x))))

library(ComplexHeatmap)

ht <- ComplexHeatmap::Heatmap(avg, 
              row_labels = as.expression(newnames),
              show_column_names = TRUE, 
              show_row_names = TRUE, 
              cluster_rows = FALSE,  # Désactiver le clustering des lignes car déjà ordonné
              cluster_columns = FALSE,
              column_names_centered = TRUE,
              show_row_dend = FALSE,
              column_dend_reorder = FALSE,
              column_names_rot = 0,
              border = "black",
              #col = viridis::inferno(256),
              row_names_gp = gpar(fontsize = 8),  
              column_names_gp = gpar(fontsize = 8, fontface = "bold"),  
              column_names_side = "top",
              heatmap_legend_param = list(
                title = "z-score",
                legend_height = unit(0.01, 'cm'),
                legend_width = unit(4.5, 'cm'),
                #legend_direction = "horizontal", 
                border = 'black'))

pdf(file.path(dir.results, "figures/Fig_3e.pdf"), useDingbats = F,  width = 10, height = 3) 
draw(ht, heatmap_legend_side="right")
dev.off()

```

## Fig 3f

```{r fig.width=6, fig.height=5}

DefaultAssay(sub) <- "RNA"
sub <- Filter_Unwanted_Genes(sub, scale_data = T)
Idents(sub) <- "ann_level_2"
sce <- as.SingleCellExperiment(sub)
reducedDim(sce, "PCA", withDimnames=TRUE) <- sub[['pca']]@cell.embeddings
reducedDim(sce, "UMAP", withDimnames=TRUE) <- sub[['umap']]@cell.embeddings
start.clus <- "Olf. HBCs"
end.clus <- c("Microvillars","Sustentaculars","mOSNs")
sce <- suppressWarnings(slingshot(sce, 
                                  thresh = 0, 
                                  reducedDim = 'UMAP', 
                                  clusterLabels = 'ann_level_2', 
                                  start.clus = start.clus,
                                  end.clus = end.clus
                                  ))
#saveRDS(sce, file.path(dir.results, "rdata/OE_slingshot.rds"))
dimred <- as.matrix(sub@reductions$umap@cell.embeddings)
clustering <- sub$ann_level_2
lineages <- SlingshotDataSet(sce)
lineages
curves <- getCurves(lineages, approx_points = NULL, thresh = 0.01, stretch = 0, allow.breaks = FALSE, shrink = 0.99)

```


```{r fig.width=6, fig.height=5}

colors <- c("Olf. HBCs"="#E41A1C","GBCs"="#F48B5A","INPs"= "#E69F00","iOSNs"="#f05b43","mOSNs"="#33b8ff", "Microvillars"="#efe13c","Sustentaculars"="#C09ACA")

pdf(file.path(dir.results, "figures/Fig_3f.pdf"), useDingbats = F, width = 6.5, height = 6) 
plot(dimred, col = colors[clustering], cex = 0.6, pch = 16, axes = FALSE, xlab = "", ylab = "")
lines(lineages,  lwd = 2, type = 'lineages', col = 'black', show.constraints = TRUE)
dev.off()

```

## Fig 3g

```{r fig.width=6, fig.height=5}

pdf(file.path(dir.results, "figures/Fig_3g.pdf"), useDingbats = F, width = 6.5, height = 6) 
plot(dimred, col = colors[clustering], cex = 0.6, pch = 16, axes = FALSE, xlab = "", ylab = "")
lines(SlingshotDataSet(curves), lwd = 2, col = "black")
dev.off()

```

```{r}

## Neuronal lineage

plot_pseudotime_umap(
  sce = sce,
  pseudotime_col = "slingPseudotime_1",
  embedding = "UMAP",
  title = "Neuronal lineage",
  color_palette = "YlGnBu",
  output_file = file.path(dir.results, "figures/Fig_3g_neuronal_lineage.pdf")
)

## Microvillar lineage

plot_pseudotime_umap(
  sce = sce,
  pseudotime_col = "slingPseudotime_2",
  embedding = "UMAP",
  title = "Microvillar lineage",
  color_palette = "YlGnBu",
  output_file = file.path(dir.results, "figures/Fig_3g_microvilar_lineage.pdf")
)

## Sustentacular lineage
plot_pseudotime_umap(
  sce = sce,
  pseudotime_col = "slingPseudotime_3",
  embedding = "UMAP",
  title = "Sustentacular lineage",
  color_palette = "YlGnBu",
  output_file = file.path(dir.results, "figures/Fig_3g_sustentacular_lineage.pdf")
)


```

## Fig 3h

```{r}

# Extract pseudotime
pseudotime_data <- as.data.frame(slingPseudotime(sce))

# Melt pseudotime data for ggplot
pseudotime_melted <- reshape2::melt(pseudotime_data, 
                          variable.name = "Lineage", 
                          value.name = "Pseudotime")

# Add cell type or cluster information
pseudotime_melted$PCW <- sce$PCW_order

# Generate density plot
p <- ggplot(pseudotime_melted, aes(x = Pseudotime, fill = PCW, color = PCW)) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  labs(title = "Differential progression in olfactory epithelium \n from PCW7 to PCW12",
       x = "Pseudotime",
       y = "Density") +
  scale_fill_manual(values = custom_colors$PCW) +
    scale_color_manual(values = custom_colors$PCW) +
    guides(fill = guide_legend(title = NULL), color = guide_legend(title = NULL)) +
  plot_theme(theme_type = "classic",legend_position = c(0.08,0.85))

print(p)
save_plot(p, filename = file.path(dir.results, "figures/OE_pseudotime_differential_progression"),formats = c("png", "pdf"),  width = 5, height = 3)    

```


## Fig 3j
### Neuronal lineage

```{r fig.width=4, fig.height=8}

gam_results <- test_pseudotime_genes_gam(
  seurat_obj = sub,
  sce_obj = sce,
  pseudotime_col = "slingPseudotime_1",
  annotation_level = "ann_level_2",
  selected_celltypes = c("Olf. HBCs", "GBCs", "INPs", "iOSNs", "mOSNs"),
  n_variable_genes = 500
)


cell_lineage1 <- c("Olf. HBCs", "GBCs", "INPs", "iOSNs", "mOSNs")
subset_sce <- sce[, sce$ann_level_2 %in% cell_lineage1 & !is.na(sce$slingPseudotime_1)]


plot_heatmap_pseudotime(
  sce = subset_sce,
  pseudotime_col = "slingPseudotime_1",
  annotation_col = "ann_level_2",
  lineage_cells = gam_results$lineage_cells,
  ptime_order = gam_results$lineage_cells[order(gam_results$ptime)],
  gene_list = head(gam_results$result$id, 50),
  celltype_order = c("Olf. HBCs", "GBCs", "INPs", "iOSNs", "mOSNs"),
  celltype_colors = c("Olf. HBCs" = "#E41A1C", "GBCs" = "#F48B5A",
                      "INPs" = "#E69F00", "iOSNs" = "#f05b43", "mOSNs" = "#33b8ff"),
  #output_pdf = file.path(dir.results, "figures/Fig_3j_heatmap_pseudotime_neuronal_lineage.pdf"),
  cluster_rows = TRUE, 
  show_annotation_legend = F,
  pdf_width = 4, pdf_height = 8,
  plot_title = "Neuronal lineage",
  font_size = 10,
  method = "complex"
)


```

### Microvillar lineage

```{r fig.width=4, fig.height=8}

gam_results <- test_pseudotime_genes_gam(
  seurat_obj = sub,
  sce_obj = sce,
  pseudotime_col = "slingPseudotime_2",
  annotation_level = "ann_level_2",
  selected_celltypes = c("Olf. HBCs",  "GBCs",  "Sustentaculars"),
  n_variable_genes = 500
)


cell_lineage <- c("Olf. HBCs", "GBCs", "Sustentaculars")
subset_sce <- sce[, sce$ann_level_2 %in% cell_lineage & !is.na(sce$slingPseudotime_2)]


plot_heatmap_pseudotime(
  sce = subset_sce,
  pseudotime_col = "slingPseudotime_2",
  annotation_col = "ann_level_2",
  lineage_cells = gam_results$lineage_cells,
  ptime_order = gam_results$lineage_cells[order(gam_results$ptime)],
  gene_list = head(gam_results$result$id, 50),
  celltype_order = c("Olf. HBCs", "GBCs", "Sustentaculars"),
  celltype_colors = c("Olf. HBCs" = "#E41A1C", "GBCs" = "#F48B5A", "Sustentaculars" = "#C09ACA"),
  #output_pdf = file.path(dir.results, "figures/Fig_3j_heatmap_pseudotime_microvillar_lineage.pdf"),
  pdf_width = 4, pdf_height = 8,
  plot_title = "Sustentacular lineage",
  cluster_rows = TRUE, show_annotation_legend = F,
  font_size = 10, method = "complex"
)


```

### Sustentacular lineage

```{r fig.width=4, fig.height=8}

gam_results <- test_pseudotime_genes_gam(
  seurat_obj = sub,
  sce_obj = sce,
  pseudotime_col = "slingPseudotime_3",
  annotation_level = "ann_level_2",
  selected_celltypes = c("Olf. HBCs",  "Microvillars"),
  n_variable_genes = 500
)


cell_lineage <- c("Olf. HBCs", "Microvillars")
subset_sce <- sce[, sce$ann_level_2 %in% cell_lineage & !is.na(sce$slingPseudotime_3)]


plot_heatmap_pseudotime(
  sce = subset_sce,
  pseudotime_col = "slingPseudotime_3",
  annotation_col = "ann_level_2",
  lineage_cells = gam_results$lineage_cells,
  ptime_order = gam_results$lineage_cells[order(gam_results$ptime)],
  gene_list = head(gam_results$result$id, 50),
  celltype_order = c("Olf. HBCs", "GBCs", "Microvillars"),
  celltype_colors = c("Olf. HBCs" = "#E41A1C", "Microvillars" = "#efe13c"),
  #output_pdf = file.path(dir.results, "figures/Fig_3j_heatmap_pseudotime_sustentacular_lineage.pdf"),
  cluster_rows = TRUE, show_annotation_legend = F,
  pdf_width = 4, pdf_height = 8,
  plot_title = "Microvilar lineage",
  font_size = 10, method = "complex"
)

```

# Figure 4
## Fig 4a

```{r fig.width=7, fig.height=25}

tf_genes <- c("EBF1","LHX2","LDB1","NEUROD2","PAX2","SIX1","BARX2","NFKBIZ")

p <- plot_tf_activity_expression(sub, tf_list = tf_genes)
print(p)
save_plot(p, filename = file.path(dir.results, "figures/Fig_4a"),formats = c("pdf"),  width = 7, height = 25)

```

## Fig 4b

```{r fig.width=8, fig.height=14}

## DecoupleR

Idents(obj) <- "ann_level_2"
sub <- Seurat_Subset(obj, cluster_id = c("Olf. HBCs", "Microvillars","Sustentaculars","GBCs","INPs","iOSNs","mOSNs"), ndims = 5, min.dist = 1)
net <- readRDS("~/Documents/projects/singlecell/hudeca_nasal/data/external/net.rds")

# Extract the normalized log-transformed counts
mat <- as.matrix(sub@assays$RNA@data)

# # Run ulm
acts <- decoupleR::run_ulm(mat = mat, 
                           net = net, 
                           .source = 'source', 
                           .target = 'target',
                           .mor='mor', 
                           minsize = 5)
saveRDS(acts, file.path(dir.results, "rdata/OE_acts.rds"))

acts <- readRDS(file.path(dir.results, "rdata/OE_acts.rds"))

# Assign TF activities as a new assay in Seurat object
sub[['tfsulm']] <- acts %>%
                    tidyr::pivot_wider(id_cols = 'source', 
                                       names_from = 'condition',
                                       values_from = 'score') %>%
                    tibble::column_to_rownames('source') %>%
                    Seurat::CreateAssayObject(.)

# Set assay and normalize data
DefaultAssay(sub) <- "tfsulm"
sub <- Seurat::ScaleData(sub)
sub@assays$tfsulm@data <- sub@assays$tfsulm@scale.data

# Define identity classes
Idents(sub) <- "ann_level_2"

# Extract activities and restructure
df <- t(as.matrix(sub@assays$tfsulm@data)) %>%
      as.data.frame() %>%
      dplyr::mutate(cluster = Seurat::Idents(sub),
                    PCW = sub$PCW) %>%  # Include PCW information
      tidyr::pivot_longer(cols = -c(cluster, PCW), 
                          names_to = "source", 
                          values_to = "score") %>%
      dplyr::group_by(cluster, PCW, source) %>%
      dplyr::summarise(mean_score = mean(score), .groups = "drop")

# Select highly variable TFs
n_tfs <- 100  # Number of top TFs to select
top_tfs <- df %>%
           dplyr::group_by(source) %>%
           dplyr::summarise(std_dev = stats::sd(mean_score)) %>%
           dplyr::arrange(-abs(std_dev)) %>%
           head(n_tfs) %>%
           dplyr::pull(source)

top_tfs <- unique(c("LHX2", "LDB1","DLX1","JUNB","PAX9","TFAP2C","FOXO1", "STAT4","ZFP560", "SAP30", "VAX2","PBX1","DLX1","DLX2","DLX5","ARX","OXT1","LHX8", top_tfs))


# Filter and reshape for heatmap
heatmap_data <- df %>%
                dplyr::filter(source %in% top_tfs) %>%
                tidyr::pivot_wider(id_cols = c("cluster", "PCW"), 
                                   names_from = "source", 
                                   values_from = "mean_score") %>%
                unite("ID", cluster, PCW, sep = "_") %>%  # Merge cluster & PCW
                tibble::column_to_rownames('ID') %>%
                as.matrix()

# Replace NA values
heatmap_data[is.na(heatmap_data)] <- 0

# Define annotation for heatmap
anno <- as.data.frame(str_split_fixed(rownames(heatmap_data), "_", 2))
rownames(anno) <- rownames(heatmap_data)
colnames(anno) <- c("Cell types", "PCW")

# Define custom order & colors
pcw_order <- c("PCW7", "PCW8", "PCW10", "PCW12")
celltype_order <- c("Olf. HBCs", "Microvillars", "Sustentaculars", "GBCs", "INPs", "iOSNs", "mOSNs")

celltype_colors <- c("Olf. HBCs" = "#E41A1C", "Microvillars" = "#efe13c", 
                     "Sustentaculars" = "#C09ACA", "GBCs" = "#F48B5A", 
                     "INPs" = "#E69F00", "iOSNs" = "#f05b43", "mOSNs" = "#33b8ff")

anno$PCW <- factor(anno$PCW, levels = pcw_order)
anno$`Cell types` <- factor(anno$`Cell types`, levels = celltype_order)

anno_colors = list("Cell types" = celltype_colors, 
                   PCW = setNames(custom_colors$PCW, pcw_order))

# Reorder matrix for heatmap
mtx <- t(heatmap_data)
mtx <- mtx[, order(factor(sub("_PCW", "", colnames(mtx)), levels = celltype_order))]
#mtx <- mtx[, order(anno$PCW, anno$`Cell types`)]
mtx <- mtx[, order(anno$`Cell types`,anno$PCW)]

# Order genes by max expression
gene_order <- order(apply(mtx, 1, which.max))
mtx <- mtx[gene_order, ]
row_labels <- lapply(rownames(mtx), function(x) bquote(italic(.(x))))

# Define color palette
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu"))
colors.use <- grDevices::colorRampPalette(colors = colors)(100)

my_breaks <- c(seq(-2, 0, length.out = ceiling(100 / 2) + 1),
               seq(0.05, 2, length.out = floor(100 / 2)))

# Combine annotations for Cell Types and PCW into a single annotation
col_annotation <- anno[, c("PCW","Cell types"), drop = FALSE]

# Plot heatmap
p <- pheatmap::pheatmap(
  mat = mtx, 
  annotation = col_annotation, 
  annotation_colors = anno_colors,
  cluster_cols = FALSE,
  cluster_rows = F,
  show_colnames = FALSE,
  fontsize = 10,
  fontsize_row = 10,
  labels_row = as.expression(row_labels),
  #color = colors.use,
  color = colorRampPalette(c("purple", "black", "yellow"))(100),
  main = "Regulon activities",
  #treeheight_col = 0,
  #treeheight_row = 0,
  border_color = "black",
  breaks = my_breaks
)

# Save output
ggsave(file.path(dir.results,"figures/Fig_4b.pdf"),
       plot = p, width = 8, height = 14, useDingbats = FALSE, dpi = 200)

```

## Fig 4c

```{r fig.width=20, fig.height=15}

tf_genes <- c("EBF1","LHX2","NEUROD2","PAX2","SIX1","SATB2","ARID1B","BARX2","IRF9","TBR1","BARX1","NFKBIZ","NSD2","NR4A2","NEUROD6")
p <- plot_tf_network(net = net, selected_sources = tf_genes)
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_4c"),formats = c("pdf"),  width = 20, height = 15)

```

# Figure 6
## Fig 6a

```{r fig.width=4, fig.height=4}

Idents(obj) <- "ann_level_2"
sub <- Seurat_Subset(obj, cluster_id = c("INPs", "iOSNs","mOSNs"), ndims = 5, min.dist = 1)
DefaultAssay(sub) <- "RNA"
ORs <- readxl::read_excel(file.path(dir.main, "data/ORs/401_ORs.xlsx"))
genes <- intersect(ORs$Gene_name, rownames(sub@assays$RNA@counts))
genes <- c(genes, "VN1R1", "VN1R2")
length(genes)

p <- plot_gene_set_expression(object = sub,  features = genes, threshold = 0.5, plot_log = T, legend_position = c(0.30,0.9), theme_type = "minimal")
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_6a"), width = 4, height = 4, formats = c("pdf"))

```

## Fig 6b
```{r fig.width=3.5, fig.height=4}

cell_levels <- c("Olf. HBCs", "Microvillars","Sustentaculars","GBCs","INPs","iOSNs","mOSNs")
sample_levels <- c("PCW7","PCW8", "PCW10","PCW12")
ORs <- readxl::read_excel(file.path(dir.main, "data/ORs/401_ORs.xlsx"))
sub <- Seurat_Subset(obj, cluster_id = c("Olf. HBCs", "Microvillars","Sustentaculars","GBCs","INPs","iOSNs","mOSNs"))

p <- plot_gene_set_transcripts(
  seurat_obj = sub,
  gene_list = ORs$Gene_name,
  group_var = "ann_level_2",
  fill_var = "PCW",
  group_levels = cell_levels,
  fill_levels = sample_levels,
  legend_position = c(0.80, 0.20),
  fill_colors = custom_colors$PCW
)

print(p)
save_plot(p, filename = file.path(dir.results, "figures/Fig_6b"),formats = c("pdf"),  width = 3.5, height = 4)    

```

## Fig_6c

```{r fig.width=4, fig.height=4.2}

sub <- Seurat_Subset(obj, cluster_id = c("iOSNs","mOSNs"), ndims = 4, min.dist = 2)

result <- analyze_OR_distribution(seurat_obj = sub, gene_list = ORs$Gene_name, meta_column = "ann_level_2", expr_cutoff = 0,
  fill_colors = c("iOSNs" = "#f05b43", "mOSNs" = "#33b8ff"), legend_pos = c(0.80, 0.90)
)
result$plot
save_plot(result$plot, filename = file.path(dir.results, "figures/Fig_5c"),formats = c("pdf"),  width = 3.5, height = 4.2)

```


## Fig 6d

```{r fig.width=5, fig.height=4}

# OSNs
Idents(obj) <- "ann_level_2"
sub <- Seurat_Subset(obj, cluster_id = c("iOSNs","mOSNs"), ndims = 4, min.dist = 2)
OR_family <- readxl::read_excel(file.path(dir.main, "data/ORs/OR_family.xlsx"))
genes <- intersect(OR_family$gene_name, rownames(sub@assays$RNA@counts))
filt <- subset(sub, features = genes)
expression_matrix <- as.matrix(filt@assays$RNA@counts)
total_counts <- rowSums(expression_matrix)
selected_genes <- names(total_counts[total_counts > 0])
filt <- subset(filt, features = selected_genes)
OR_family <- subset(OR_family, gene_name %in% rownames(filt))

df <- OR_family %>% 
  dplyr::group_by(family, class) %>% 
  dplyr::summarise(count = n())

df$family <- factor(df$family)
df$class <- factor(df$class)

p <- ggplot(df, aes(x = family, y = count, fill = class)) +
  geom_col(stat = "identity", position=position_dodge()) +
  labs(x = "OR families", y = "number of ORs", title = "") +
  guides(fill=guide_legend(title="Class")) +
   plot_theme(theme_type = "minimal", 
               legend_position = c(0.9, 0.9), 
               font_size = 10, x_angle = 0) +
  scale_fill_manual(values = c("#BCB82C","#20679B")) 

print(p)
save_plot(p, filename = file.path(dir.results, "figures/Fig_6d"),formats = c("pdf"),  width = 5, height = 4)    

OR_family <- OR_family[,c("gene_name","family","class")]
write.table(OR_family, file.path(dir.results, "tables/OR_family.csv"), sep = ";", row.names = F, quote = F)

```


# Supplementary figures

## Extended data figures 1

## Extended data figures 2

### Fig S2b

```{r fig.width=16, fig.height=16, warning=F, message=F}

p <- Feature_Plot(obj, features = c("DCN","SGCZ","ITGA8",  # Mesenchyme
                                    "RUNX2","ANO5", # osteoblasts
                                    "ACAN","COL9A1", # cartilage
                                    "PDGFRB","TRPC6", # stromal
                                    "GRHL2", "TP63","CFTR", # epithelial
                                    "DCX", "GAP43","TUBB3", # neuron
                                    "PECAM1", "VWF", # endothelial
                                    "PROX1","STAB2", # lymphatic
                                    "MYO18B","MEGF10", # myocytes
                                    "CDH19", "SOX10", # glial
                                    "PTPRC","DOCK2" # immune
                               ), no_axes = T, order = F, merge_legend = T)
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_S2b"),  width = 16, height = 16, formats = c("pdf"))

```

## Extended data figures 3


## Extended data figures 4
### Fig S4a

```{r}

# Load Seurat's built-in cell cycle gene lists
s.genes <- Seurat::cc.genes$s.genes
g2m.genes <- Seurat::cc.genes$g2m.genes
postm.genes <- c("DCX", "TUBB3", "MAP2", "RBFOX3", "NEUROD1", "SOX9", "ELAVL4") 

# Run Cell Cycle Scoring
obj <- CellCycleScoring(object = obj, 
                        s.features = s.genes, 
                        g2m.features = g2m.genes, 
                        set.ident = TRUE)

# Calculate PostM Score based on PostM gene expression
obj$PostM.Score <- Seurat::FetchData(obj, vars = postm.genes) %>% rowMeans()

# Define PostM threshold (75th percentile of PostM score)
threshold <- quantile(obj$PostM.Score, 0.75)  

# Identify Non-Cycling Cells (Low UMI & low cycle scores)
nc_threshold <- 1000  # Adjust as needed
obj$NonCycling <- obj$S.Score < 0.2 & obj$G2M.Score < 0.2 & obj$nFeature_RNA < nc_threshold

# Update Cell Cycle Phases
obj$Phase <- ifelse(obj$Phase == "G1" & obj$PostM.Score > threshold, "PostM", obj$Phase)
obj$Phase <- ifelse(obj$NonCycling, "Non-Cycling", obj$Phase)  # Assign Non-Cycling

# Convert to factor with correct order
obj$Phase <- factor(obj$Phase, levels = c("G1", "S", "G2M", "PostM","Non-Cycling"))

p <- DimPlot(obj, group.by = "Phase", cols = custom_colors$phase)
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_S4a"),  width = 7, height = 5, formats = c("pdf"))

```

### Fig S4b

```{r fig.width=3, fig.height=3}

p <- Plot_Cell_Fractions(obj, x.col = "age", fill.col = "Phase",legend.title = "Phases", line.size = 1,
                  custom.colors = custom_colors$phase, x.title = "Age (PCW)", plot.type = "line", legend.position = "none")
p
save_plot(p, filename = file.path(dir.results, "figures/Fig_S4b"),formats = c("pdf"),  width = 3, height = 3)

```

### Fig S4c

```{r fig.width=14, fig.height=4, warning=F}

p <- FeaturePlot(obj, features = c("TOP2A","MKI67","CENPF"), ncol = 3)
p
save_plot(p, filename = file.path(dir.results, "figures/Fid_S4c.1"),  width = 14, height = 4, formats = c("pdf"))

```

```{r fig.width=14, fig.height=4, warning=F}

p <- FeaturePlot(obj, features = c("S.Score","G2M.Score","PostM.Score"))
p
save_plot(p, filename = file.path(dir.results, "figures/Fid_S4c.2"),  width = 14, height = 4, formats = c("pdf"))

```



### Fig S4d

```{r fig.width=8, fig.height=6, warning=F}

p <- StackedVlnPlot(object = obj, group.by = "ann_level_2_order", cols = custom_colors$ann_level_2, features = c("S.Score","G2M.Score","PostM.Score"))
p
save_plot(p, file.path(dir.results, "figures/Fig_S4d"), formats = c("pdf"),width = 8, height = 6)

```

### Fig S4e


```{r fig.width=8, fig.height=3, warning=F}

p <- plot_props(data = obj, x = "ann_level_2_order", y = "Phase", plot_type = "bar", prop = T, stack = T,font_size = 10,
           prop_multi = F, theme_type = "minimal", legend = F, colors = custom_colors$phase, x_title = "", x_angle = 45)

p
save_plot(p, filename = file.path(dir.results, "figures/Fig_S4e"),formats = c("pdf"),  width = 8, height = 3)

```


## Extended data figures 5

